---
geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm"
header-includes:
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
output: pdf_document
#output: html_document
---

```{r setup0, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\bcenter
# ST230 Lab Assignment 1 R Worksheet
\ecenter

Before each lab, an R worksheet like this one will be available on MyLearningSpace. Each worksheet will demonstrate R code that will be used in the coming lab.  
**It's important that you work through the worksheet before the lab.** 

Worksheets will be available as R markdown files (file extension *.Rmd) and are to be opened in RStudio. Plain text can be input in any of the white spaces. R code may be entered in any of the grey spaces between the commands '''{r} and '''. These areas are called code chunks. 

```{r comments}
# This is inside a code chunk.
# Furthermore, the '#' is used for comments and turns the text green. 
# Comments are not executed when running the contents of a code chunk.
# Comments are important - read them!
```

To execute a code chunk, push the run button (which looks like a play button) in the top right of the code chunk. Alternatively, pressing "Ctrl+Shift+Enter" when the cursor is in the code chunk will execute the code within.

The necessary code chunks will generally be given. If you would like to insert an additional code chunk, you can simply type '''{r} and ''' (with a new line in between) or use the quick key combination "Ctrl"+"Alt"+"I".

## Defining Variables

- To assign a scalar (i.e. a number) or a vector (i.e. a list of numbers) a variable name in R use "<-" (less than symbol followed by minus symbol)
- Vectors are defined using the `c()` command

### Example 1

Execute the code chunk below. Notice that when X, Y and S are defined the results are only displayed by typing "X", "Y" or "S" afterwards.

```{r}
# Assign the value 2 to X:
X <- 2
# Display X
X 
# Define Y as a vector containing the values 1,2,3,4
Y <- c(1,2,3,4) 
# Display Y
Y
# Values of a vector need not be numbers. 
# For example, S is defined as the sample space from example 1.1 (Family Planning) in the lecture notes:
S <- c("MMM","FFF","MFM","FMM","MMF","MFF","FMF","FFM")
# Display S
S
```

### Question: Sample Space and Events

Consider the following example:  
Suppose that vehicles taking a particular freeway exit can turn right (R), turn left (L), or go straight (S). Consider observing the direction for each of three successive vehicles.

 a. List all outcomes in the event $A$ that all three vehicles go in the same direction.
 b. List all outcomes in the event $B$ that all three vehicles take different directions.
 c. List all outcomes in the event $C$ that exactly two of the three vehicles turn right.
 d. List all outcomes in the event $D$ that exactly two vehicles go in the same direction.

Define each of the events as vectors in the code chunk below. Use the given names.  
For example, part a would be `A <- c("LLL","RRR","SSS")`.

```{r}
# Replace "NULL" with the proper code
A <- c("LLL", "RRR", "SSS")
B <- c("LRS", "LSR", "RLS", "RSL", "SLR", "SRL")
C <- c("RRL", "RRS", "RLR", "RSR", "LRR", "SRR")
D <- c("LLR", "LLS", "LRL", "LSL", "RLL", "SLL","RRL", "RRS", "RLR", "RSR", "LRR", "SRR","SSR", "SSL", "SRS", "SLS", "LSS", "RSS")

```

R is able to preform basic set operations: 
 
 - `union(x,y)` produces the union of sets `x` and `y`
 - `intersect(x,y)` produces the intersection of sets `x` and `y`
 - `setdiff(x,y)` produces the values in set `x` that are not in set `y`
 
Use the events defined in the previous code chunk to define event $E$ as $C$ union $D$ and event $G$ as $C$ intersect $D$ in the following code chunk.  
For example, $E$ would be defined by: `E <- union(C,D)`.

```{r}
# Replace "NULL" with the proper code
E <- union(C, D)
G <- intersect(C, D)

```

## Basic Functions

As with many programming languages, R has many functions available for use. So far this worksheet has made use of the functions `c()`, `union()` and `intersect()`. To see all the details of a function simply type "?" before the function name and execute the command.

### Question: Help `?`

If you want help on the `c()` function for example, simply type `?c` and execute the code chunk.  
Type `?c` in the following code chunk and run it. The documentation should appear in the bottom right corner of this window.

```{r, echo=FALSE}
?C

```

Read the description. What could the letter 'c' stand for in this function?  
**Type your answer below:**  
Sets the "contrasts" attribute for the factor.


## Permutations and Combinations

Recall that the number of **permutations** (order matters) of $r$ objects selected from a set of $n>r$ distinct objects is:
$$_nP_r = n(n-1)\cdots(n-r-1) = \dfrac{n!}{(n-r)!}$$
Recall that the number of **combinations** (order doesn't matter) of $r$ objects selected from a set of $n>r$ distinct objects is :
$$_nC_r = \binom{n}{r} = \dfrac{n(n-1)\cdots(n-r-1)}{r!} = \dfrac{n!}{r!(n-r)!}$$
R comes with a `factorial()` function, `factorial(n)` will calculate $n!$. This can be used to calculate permutations and combinations in R.

### Example 2

From an alphabet consisting of 10 digits, 26 lower-case and 26 upper-case letters, how many different 8-character passwords can one create? 

There are 10 + 26 + 26 = 62 different characters. The number of different 8-character passwords is: $$\dfrac{62!}{(62-8)!}$$

In R the calculation would be:

```{r}
# Store the number of passwords in a variable called "num_of_8_char_pass"
num_of_8_char_pass <- factorial(62)/factorial(62-8) # or: factorial(62)/factorial(54)
# Display the number of described passwords:
num_of_8_char_pass
```

### Question: Counting

Suppose a science council has one student representative from each of the following departments: chemistry, mathematics, biology, physics, and psychology. In how many ways can:

 a. Calculate the number of ways a president and vice president can be appointed from the science council. Store the result in a variable named `total_pres`.
 b. Calculate the number of ways a president, vice president and treasurer can be appointed from the science council. Store the result in a variable named `total_office`.
 c. Calculate the number of ways two members from the science council could be chosen to serve on the Dean's Council. Store the result in a variable named `total_dean`.

Preform each of the calculations in the code chunk that follows. **Use the specified variable names.**

```{r}
# Replace "NULL" with the proper code
total_pres <- factorial(5) / factorial(5-2) # 2 roles
total_office <- factorial(5) / factorial(5-3) # 3 roles
total_dean <- factorial(5) / (factorial(2) * factorial(3)) # combinations of 2


```

### Question: Rolling Two Dice

Creating R functions (which can be used for simulations) is beyond the scope of this course. You will, however, be asked to run functions (and / or simulations) and analyze results.

An example of an R function can be found in the following code chunk. Read the comments carefully to understand the code.

```{r}
# This code is a function that simulates n rolls of two fair multi-sided dice. 
# Each roll, the dice results are added together and stored in a vector called results.
dice_roll <- function(sides=6,rolls=1) {
  # initialize the results vector (empty to begin)
  results <- c()
  # initiate a for loop to run 'rolls' times
  for (i in 1:rolls) {
    # simulate the rolling of two die (randomly select two values from {1,2,3,..,'sides'})
    roll <- sample(1:sides,size=2,replace=TRUE)
    # sum the results of the roll (result will be a number between 2 and 2*sides) and put the total in the results vector
    results <- c(results,sum(roll))
  }
  # print the results
  return(results)
}
# Execute this code chunk before proceeding.
```

Once the previous code chunk has been run, the function `dice_roll()` is defined in our environment and can be used in subsequent code chunks. Type `dice_roll(sides = 8, rolls = 1)` and `dice_roll(sides = 9, rolls = 10)` in the following code chunk and run it to simulate one roll and ten rolls. Take a moment to understand the outcome.

```{r}
dice_roll(sides = 8, rolls = 1)
dice_roll(sides = 9, rolls = 10)

```

Complete the following tasks in the next code chunk and execute it:

 1. Use the `dice_roll()` function to simulate 500 rolls of two 9-sided dice. Store the results in `rolls_500`.
 2. `table(rolls_500)` will provide a summary of the outcomes stored in `rolls_500`; view the summary.
 3. `barplot(table(rolls_500))` will generate a bar chart of the outcomes stored in `rolls_500`; view the barplot.

```{r}
# Replace "NULL" with the proper code
# Use rolls_500 to complete the stated tasks:
rolls_500 <- dice_roll(sides = 9, rolls = 500)
table(rolls_500)
barplot(table(rolls_500))


```

### Submission Instructions

1. Save this file. (Quick key combo for that is "control" (or "command" on a Mac) and "s")
2. Run the following code chunk to produce a pdf.
3. Both of these files are likely in the "Downloads" folder (unless you moved them). Check the contents of each file and upload both the Rmd and pdf files to Gradescope.

``` {r, echo=FALSE ,eval=FALSE}
rmarkdown::render(rstudioapi::getSourceEditorContext()$path,output_format = "pdf_document")
```
