---
geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm"
header-includes:
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
output: pdf_document
#output: html_document
---

```{r setup0, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\bcenter
# ST230 Lab Assignment 3 Prep
\ecenter

$>$bworksheet
Recall: to execute a code chunk, push the run button (which looks like a play button) in the top right of the code chunk. Alternatively, pressing "Ctrl+Shift+Enter" when the cursor is in the code chunk will execute the code within.

Code chunks will always start with `'''{r}` and end with `'''`.  If necessary, to insert a code chunk you can click on "Insert" along the tool bar and choose R or "Ctrl+Alt+I".  For the most part, code chunks will be given and you won't need to insert them into the file.
$>$eworksheet

### Discrete Distributions

The majority of the pmfs for discrete distributions are quite simple to work with.  They can, however, be quite cumbersome when you are required to perform multiple calculations.

Within R there are many commands associated with the discrete distributions discussed in lecture (see MyLS lab notes for a refresher, if necessary).  Below is a chart containing each of the distributions discussed so far in the course along with how they are referenced in R:

[Note: See lab notes for definitions of n, p, etc.]

Distribution Name:            R:

Binomial                      binom(x,n,p)

Hypergeometric                hyper(x,a,N-a,n)

Geometric                     geom(x-1,p)

Negative Binomial             nbinom(x-r,r,p)

Poisson                       pois(x,lambda)

To use any of these distributions in R, you must precede them with:
 - 'd' for the probability mass function P(X=x) 
 - 'p' for the cumulative distribution P(X<=x)
 - 'q' to find quantiles (percentiles)
 - 'r' to generate a random sample

Consider the following example code:

```{r}
# Assume X ~ Binomial(n=15,p=1/6)
# Calculates P(X = 3) 
dbinom(3,15,1/6)
# NOTE: the first input of dbinom() can be a vector of numbers. 
# The following returns a vector of probabilities: c(P(X = 1),P(X = 2),P(X = 3))
dbinom(c(1,2,3),15,1/6)
# Calculates P(X < 4)   NOTE, this is X less than or equal to 3
pbinom(3,15,1/6)
# Determines x so that P(X < x) is approximately 0.1  (X is discrete)
qbinom(0.1,15,1/6)
# Runs 15 trials 10 times and records the number of successes each time.
rbinom(10,15,1/6)
```

#### Binomial Distribution Exercise

Let $X$ represent the number of heads that are flipped in five coin flips. In this case, $X$ ~ Binomial(n=5,p=0.5). Complete each of the following tasks in the next code chunk: 
 (a) Use the `c()` function to define `X` as the possible outcomes of this experiment.
 (b) Use `dbinom()` to define `prob` as the probability distribution of `X`. That is, `prob` is the values of the *pmf* of $X$.
 (c) Use `pbinom()` to define `prob_3` as the probability of getting *more than* 3 heads in this experiment.
 (d) Use `rbinom()` to define `sim_disc` as the value of $X$ after simulating this experiment 1000 times.
 (e) Apply the `mean()` function to determine the average of `sim_disc`. How does this compare to $E(X)$?
 
```{r, five.coins}
# (a) NOTE: In R, `c(3,4,5,6,7,8)` and `3:8` will produce the same result. 
#           `a:b` can be used to produce integers from `a` to `b`.
X <- 0:5
# (b)
prob <- dbinom(X, size = 5, prob = 0.5)
# (c)
prob_3 <- 1 - pbinom(3, size =5, prob = 0.5)
# (d)
sim_disc <- rbinom(1000, size = 5, prob = 0.5)
# (e)
mean_sim_disc <- mean(sim_disc)

```

#### Hypergeometric Distribution Exercise

A pharmaceutical company inspects a simple random sample (SRS) of 10 empty plastic containers from a shipment of 10,000 containers. The containers are examined for traces of benzene, a common chemical solvent but also a known human carcinogen. Suppose that (unknown to the company) 10\% of the containers in the shipment contain traces of benzene. Let $X$ be the number of containers contaminated with benzene in the sample.

Have R calculate the probability that there is no more than one contaminated container in the sample. Assign the value to `hyper_prob`.

```{r, hyper}
hyper_prob <- phyper(1, m =1000, n =9000, k =10)

```

### Continuous Distributions

Within R there are many commands associated with the continuous distributions discussed in lecture (see MyLS lab notes for a refresher, if necessary).  Below is a chart containing each of the distributions discussed so far in the course along with how they are referenced in R:

[Note: See lab notes for definitions of n, p, etc.]

Distribution Name:            R:

Uniform                       unif(x,alpha,beta)

Normal                        norm(x,mu,sigma)

To use any of these distributions in R, you must precede them with:
 - 'd' for the probability density function f(x) 
 - 'p' for the cumulative distribution P(X<=x)
 - 'q' to find quantiles (percentiles)
 - 'r' to generate a random sample

Consider the following example code:

```{r}
# Assume X ~ Normal(mu = 0,sigma = 1)
# dnorm(1,0,1) calculates f(1); where f(x) is the normal pdf. 
# This gives the height of the curve and is not a probability. 
dnorm(1,0,1)
# pnorm(1,0,1) calculates P(X < 1)   NOTE, this is X less than or equal to 1
pnorm(1,0,1)
# qnorm(0.1,0,1) determines x so that P(X < x) = 0.1
qnorm(0.1,0,1)
# rnorm(10,0,1) generates 10 random samples from the indicated normal distribution.
rnorm(10,0,1)
```

#### Normal Distribution Exercise

The distribution of heights of young females age 18 to 24 is approximately Normal, with mean mu = 64.5 inches and standard deviation sigma = 2.5 inches.
Let $X$ be the height of a randomly selected female (age 18 to 24). $X$ ~ Normal(mu = 64.5,sigma = 2.5). Complete each of the following tasks in the next code chunk: 
 (a) Use `pnorm()` to define `prob_60_64` as the probability that a randomly selected female is between 60 inches and 64 inches tall.
 (b) Use `pnorm()` to define `prob_67` as the probability that a randomly selected female is taller than 67 inches.
 (c) Use `qnorm()` to define `p_90` determine the 90th percentile of $X$.
 (d) Use `rnorm()` to define `sim_cts` as a vector of the heights of 1000 randomly selected females.
 (d) Apply the `mean()` function to determine the average of `sim_cts`. How does this compare to $E(X)$?
 
```{r, heights}
# (a) 
prob_60_64 <- pnorm(64, mean = 64.5, sd = 2.5) - pnorm(60, mean = 64.5, sd = 2.5)
# (b)
prob_67 <- 1 - pnorm(67, mean = 64.5, sd = 2.5)
# (c)
p_90 <- qnorm(0.9, mean = 64.5, sd = 2.5)
# (d)
sim_cts <- rnorm(1000, mean =64.5, sd = 2.5)
# (e)
mean_sim_cts <- mean(sim_cts)

```

### Submission Instructions

1. Save this file. (Quick key combo for that is "control" (or "command" on a Mac) and "s")
2. Run the following code chunk to produce a pdf.
3. Both of these files are likely in the "Downloads" folder (unless you moved them). Check the contents of each file and upload both the Rmd and pdf files to Gradescope.

``` {r, echo=FALSE ,eval=FALSE}
rmarkdown::render(rstudioapi::getSourceEditorContext()$path,output_format = "pdf_document")
```
